Intro to Predictive Modeling

0. Complete

```{r}
# Load packages


library(FNN)
library(ranger)
library(rms)
library(nnet)
library(ggplot2)
library(dplyr)

```

```{r}

# Load data
abalone <- read.csv(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data",
  header = FALSE,
  col.names = c(
    "sex", "length", "diameter", "height",
    "whole_wt", "shucked_wt",
    "viscera_wt", "shell_wt", "rings"
  )
)

abalone$sex <- factor(abalone$sex)

# 70/30 split
train_idx <- sample(1:nrow(abalone), size = 0.70 * nrow(abalone))

train <- abalone[train_idx, ]
test <- abalone[-train_idx, ]
```



1. KNN

```{r}
# KNN
x_train <- model.matrix(rings ~ ., data = train)[, -1]
x_test <- model.matrix(rings ~ ., data = test)[, -1]

x_train_scaled <- scale(x_train)

x_test_scaled <- scale(
  x_test,
  center = attr(x_train_scaled, "scaled:center"),
  scale = attr(x_train_scaled, "scaled:scale")
)

pred_knn <- FNN::knn.reg(
  train = x_train_scaled,
  test = x_test_scaled,
  y = train$rings,
  k = 10
)$pred
```

1.1. Random Forest
```{r}
rf_model <- ranger(
  rings ~ .,
  data = train
)

pred_rf <- predict(rf_model, data = test)$predictions
```

1.2. Ordinal Regression Model
```{r}
dd <- datadist(train)
options(datadist = "dd")

orm_model <- orm(
  rings ~ sex +
    rcs(length, 4) +
    rcs(diameter, 4) +
    rcs(height, 4) +
    rcs(whole_wt, 4) +
    rcs(shucked_wt, 4) +
    rcs(viscera_wt, 4) +
    rcs(shell_wt, 4),
  data = train
)

pred_orm <- predict(orm_model, newdata = test, type = "mean")
```

1.3 Linear Regression Model

```{r}
ols_model <- ols(
  rings ~ sex +
    rcs(length, 4) +
    rcs(diameter, 4) +
    rcs(height, 4) +
    rcs(whole_wt, 4) +
    rcs(shucked_wt, 4) +
    rcs(viscera_wt, 4) +
    rcs(shell_wt, 4),
  data = train
)

pred_ols <- predict(ols_model, newdata = test)
```

1.4 Neural Network Model
```{r}
nn_model <- nnet(
  rings ~ .,
  data = train,
  size = 5,
  linout = TRUE,
  trace = FALSE,
  maxit = 500
)

pred_nn <- as.vector(predict(nn_model, newdata = test))
```

2. 

```{r}
cal_plot <- function(obs, pred, title) {
  data.frame(
    observed = obs,
    predicted = pred
  ) |>
    ggplot(aes(x = predicted, y = observed)) +
    geom_point(alpha = 0.35) +
    geom_smooth(method = "loess", se = FALSE) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    labs(
      title = title,
      x = "Predicted Rings",
      y = "Observed Rings"
    )
}
```

```{r}
cal_plot(test$rings, pred_knn, "KNN Calibration Plot")
cal_plot(test$rings, pred_rf, "Random Forest Calibration Plot")
cal_plot(test$rings, pred_orm, "Ordinal Regression Calibration Plot")
cal_plot(test$rings, pred_ols, "Linear Regression with RCS Calibration Plot")
cal_plot(test$rings, pred_nn, "Neural Network Calibration Plot")
```

```{r}
mse_results <- data.frame(
  Model = c(
    "KNN",
    "Random Forest",
    "Ordinal Regression with RCS",
    "Linear Regression with RCS",
    "Neural Network"
  ),
  MSE = c(
    mean((test$rings - pred_knn)^2),
    mean((test$rings - pred_rf)^2),
    mean((test$rings - pred_orm)^2),
    mean((test$rings - pred_ols)^2),
    mean((test$rings - pred_nn)^2)
  )
)

mse_results |> arrange(MSE)
plot(
  pred_nn,
  test$rings,
  main = "Neural Network Calibration Plot",
  xlab = "Predicted",
  ylab = "Observed"
)

abline(0, 1)
lines(lowess(pred_nn, test$rings))
```

3. 
 The best model on the Abalone data was the neural network because it had the lowest test-set MSE, 4.243645. I am defining "best" as the model that makes the most accurate predictions on new test data. Since MSE measures the average squared difference between the predicted and observed number of rings, the lowest MSE means the model made the smallest prediction errors overall.

The neural network likely performed best because it can capture nonlinear patterns between the abalone measurements and the number of rings. However, "best" can depend on the goal. If the main goal is interpretability, linear regression with restricted cubic splines may be preferred because it had a similar MSE and is easier to explain. Random forest and ordinal regression also performed reasonably well, while KNN had the highest MSE in this run.

Overall, the model choice depends on whether the priority is prediction accuracy, calibration, interpretability, or simplicity. For this assignment, I would choose the neural network because I am using test-set MSE as my main definition of best.