---
title: "Module 8 Homework"
author: "Liane Nicole"
format: html
---

```{r}
#| message: false
#| warning: false

library(glmnet)
library(knitr)

set.seed(6021)

# Instructor's workaround
d1 <- readRDS("riboflavin.rds")

X <- as.matrix(d1$x)
y <- d1$y

n <- nrow(X)
p <- ncol(X)

# Use the same folds for all models
fold_id <- sample(rep(1:10, length.out = n))
```

# Problem 0

I created `HW8.qmd` and added it to `_quarto.yml`.



# Problem 1

Because there are `r p` predictors but only `r n` observations, \(p > n\).
Therefore, ordinary least squares cannot produce a unique solution.

```{r}
ols_data <- as.data.frame(X)
ols_data$y <- y

ols_fit <- lm(y ~ ., data = ols_data)

na_count <- sum(is.na(coef(ols_fit)))
na_count
```

R returns `r na_count` coefficients as `NA` because the predictors are
linearly dependent. There are more predictors than observations, so the model
matrix does not have full rank.

# Problem 2

```{r}
ridge_cv <- cv.glmnet(
  X,
  y,
  alpha = 0,
  foldid = fold_id
)

ridge_cv$lambda.min
ridge_cv$lambda.1se
```

\[
\lambda_{\min} = \text{`r round(ridge_cv$lambda.min, 6)`}
\]

\[
\lambda_{1se} = \text{`r round(ridge_cv$lambda.1se, 6)`}
\]

```{r}
plot(ridge_cv)
abline(v = log(ridge_cv$lambda.min), lty = 2)
abline(v = log(ridge_cv$lambda.1se), lty = 3)
```

As \(\lambda\) increases, CV-MSE usually decreases at first, reaches a
minimum, and then increases when the model becomes too strongly shrunk.

# Problem 3

```{r}
ridge_fit <- glmnet(X, y, alpha = 0)

plot(
  ridge_fit,
  xvar = "lambda",
  xlab = expression(log(lambda)),
  ylab = "Coefficient value"
)
```

As \(\lambda\) increases, the ridge coefficients shrink toward zero. Ridge
does not usually make coefficients exactly zero.

# Problem 4

```{r}
lasso_cv <- cv.glmnet(
  X,
  y,
  alpha = 1,
  foldid = fold_id
)

lasso_coef_1se <- coef(lasso_cv, s = "lambda.1se")

lasso_selected_1se <- sum(lasso_coef_1se[-1] != 0)

lasso_selected_1se
```

At \(\lambda_{1se}\), the lasso keeps
**`r lasso_selected_1se` gene-expression features**.

```{r}
lasso_fit <- glmnet(X, y, alpha = 1)

plot(
  lasso_fit,
  xvar = "lambda",
  xlab = expression(log(lambda)),
  ylab = "Coefficient value"
)
```

```{r}
plot(lasso_cv)
abline(v = log(lasso_cv$lambda.min), lty = 2)
abline(v = log(lasso_cv$lambda.1se), lty = 3)
```

Unlike ridge, lasso can make some coefficients exactly zero.

# Problem 5

```{r}
alphas <- c(0.25, 0.50, 0.75)

enet_results <- data.frame(
  alpha = alphas,
  lambda_min = NA,
  min_CV_MSE = NA,
  selected_features = NA
)

enet_fits <- vector("list", length(alphas))

for (i in seq_along(alphas)) {
  enet_fits[[i]] <- cv.glmnet(
    X,
    y,
    alpha = alphas[i],
    foldid = fold_id
  )

  enet_results$lambda_min[i] <- enet_fits[[i]]$lambda.min
  enet_results$min_CV_MSE[i] <- min(enet_fits[[i]]$cvm)

  enet_coef <- coef(enet_fits[[i]], s = "lambda.min")
  enet_results$selected_features[i] <- sum(enet_coef[-1] != 0)
}

kable(enet_results, digits = 6)
```

```{r}
best_enet_row <- which.min(enet_results$min_CV_MSE)

best_alpha <- enet_results$alpha[best_enet_row]
best_enet_mse <- enet_results$min_CV_MSE[best_enet_row]
best_enet_selected <- enet_results$selected_features[best_enet_row]

lasso_coef_min <- coef(lasso_cv, s = "lambda.min")
lasso_selected_min <- sum(lasso_coef_min[-1] != 0)
```

The elastic-net model with the lowest CV-MSE uses
\(\alpha =\) **`r best_alpha`**.

It selects `r best_enet_selected` features at \(\lambda_{\min}\), while lasso
selects `r lasso_selected_min` features at \(\lambda_{\min}\).

Elastic net may keep more correlated predictors than lasso because it combines
the ridge and lasso penalties.

# Problem 6

I used manual principal component regression with 10-fold cross-validation.

```{r}
pcr_mse <- rep(0, 20)

for (k in 1:20) {
  fold_errors <- c()

  for (fold in 1:10) {
    train <- fold_id != fold
    test <- fold_id == fold

    x_train <- scale(X[train, ])
    train_center <- attr(x_train, "scaled:center")
    train_scale <- attr(x_train, "scaled:scale")

    train_scale[train_scale == 0] <- 1

    x_test <- scale(
      X[test, ],
      center = train_center,
      scale = train_scale
    )

    pca <- prcomp(x_train, center = FALSE, scale. = FALSE)

    train_scores <- pca$x[, 1:k, drop = FALSE]
    test_scores <- x_test %*% pca$rotation[, 1:k, drop = FALSE]

    pcr_fit <- lm(y[train] ~ train_scores)

    prediction_data <- as.data.frame(test_scores)
    names(prediction_data) <- paste0("PC", 1:k)

    fit_data <- as.data.frame(train_scores)
    names(fit_data) <- paste0("PC", 1:k)

    pcr_fit <- lm(y[train] ~ ., data = fit_data)
    predictions <- predict(pcr_fit, newdata = prediction_data)

    fold_errors <- c(fold_errors, (y[test] - predictions)^2)
  }

  pcr_mse[k] <- mean(fold_errors)
}

optimal_k <- which.min(pcr_mse)
optimal_k
```

```{r}
plot(
  1:20,
  pcr_mse,
  type = "b",
  xlab = "Number of components",
  ylab = "CV-MSE"
)

abline(v = optimal_k, lty = 2)
```

The optimal number of principal components is
**`r optimal_k`**.

## Problem 6a

```{r}
X_scaled <- scale(X)

pca_full <- prcomp(
  X_scaled,
  center = FALSE,
  scale. = FALSE
)

scores_full <- pca_full$x[, 1:optimal_k, drop = FALSE]

pcr_full <- lm(y ~ scores_full)

gamma_hat <- coef(pcr_full)[-1]

beta_scaled <- pca_full$rotation[, 1:optimal_k, drop = FALSE] %*%
  gamma_hat

x_center <- attr(X_scaled, "scaled:center")
x_scale <- attr(X_scaled, "scaled:scale")

beta_pcr <- as.vector(beta_scaled) / x_scale

intercept_pcr <- coef(pcr_full)[1] -
  sum(x_center * beta_pcr)

length(beta_pcr)
```

The back-transformed PCR coefficient vector has
`r length(beta_pcr)` coefficients.

## Problem 6b

```{r}
prediction_1 <- intercept_pcr + sum(X[1, ] * beta_pcr)

prediction_1
```

The predicted value for the first observation is
**`r round(prediction_1, 6)`**.

## Problem 6c

All 4,088 gene-expression values are needed because every principal component
is created from a weighted combination of all original predictors.

# Problem 7

```{r}
ridge_mse <- min(ridge_cv$cvm)
lasso_mse <- min(lasso_cv$cvm)
enet_mse <- best_enet_mse
pcr_best_mse <- min(pcr_mse)

comparison <- data.frame(
  Method = c("Ridge", "Lasso", "Elastic net", "PCR"),
  CV_MSE = c(
    ridge_mse,
    lasso_mse,
    enet_mse,
    pcr_best_mse
  )
)

kable(comparison, digits = 6)
```

```{r}
best_method <- comparison$Method[
  which.min(comparison$CV_MSE)
]

best_method
```

The method with the lowest CV-MSE is
**`r best_method`**.

Ridge is often expected to work well when many correlated predictors have
small effects. Lasso is often better when only a small number of predictors
are truly important. Comparing the ridge and lasso CV-MSE values above shows
which pattern fits this dataset better.
